---
title: "Code sample"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidyr)
library(readxl)
library(gridExtra)
```

The following project is an exploratory analysis examining trends in the prevalence of EEOP, Enlarged External Occipital Protuberance or bony projections, in young people.

```{r, include=FALSE}
posture = 
  read_excel("./data/p8105_mtp_data.xlsx", skip = 8) %>% 
  janitor::clean_names() %>% 
mutate(
    eop_size_mm = replace_na(eop_size_mm,0),
    sex = factor(sex, c(0,1), c("female","male")),
    age_group = factor(age_group, c(2,3,4,5,6,7,8), c("18-30","31-40","41-50","51-60","60+","60+","60+"), ordered = TRUE),  
    eop_size = factor(eop_size, c("0", "1", "2", "3", "4", "5"), c("0-5", "5-10", "10-15", "15-20", "20-25", "25+"), ordered = TRUE),
    eop_visibility_classification = factor(eop_visibility_classification, c("0","1","2"), c("eop size 0 mm", "0 < eop size mm <=5","eop size >= 5 mm"), ordered = TRUE), 
    eop_shape = factor(eop_shape, c("1","2","3","4","5"), c("1","2","3","4","5"), ordered = TRUE),
    fhp_category = factor(fhp_category)) %>% 
filter(age_group != "1")
posture
```

### Data Summary

There are `r nrow(posture)` participants and `r ncol(posture)` variables in the dataset. Significant variables include: `eom_size_mm`, which indicates the size of the protrusion, and `fhp_size_mm`, which indicates the extent of the participants' forward head protraction, a measure that is potentially linked to EOP. The largest proportions of participants are between 18-30 and over 60. Sex is fairly evenly distributed. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
age_sex=
posture %>% 
  group_by(age_group, sex) %>% 
  summarize (n_obs = n()) %>% 
  pivot_wider(
    names_from = sex,
    values_from = n_obs
  )

knitr::kable(age_sex)

posture %>% 
 ggplot(aes(x = age, fill = sex, color=sex)) + geom_histogram(color="black")
```

EEOP is defined as an EOP larger than 10mm. We can assess the prevalence of enlarged EOPs in the data by filtering `eop_size_mm` for EOPs larger than 10mm, finding an EEOP rate of 32.16% in the sample.

```{r, echo=FALSE}
eeop_rate =
  posture %>% 
  mutate(eeop = as.numeric(ifelse(eop_size_mm > 10, "1", "0"))) %>% 
  summarize (eeop_rate = mean(eeop))

knitr::kable(eeop_rate)
```

In data cleaning, irregularities were found in categorical variables `eop_size_mm` and `fhp_category` where data does not adhere to the defined categories. There are missing data in `eop_size_mm`, `eop_shape`, and `fhp_size_mm`.

### Trends in age and sex

Our analysis found that men consistently had higher FHP measurements than women, and that size of FHP increases with age. For reference, the mean female FHP size was `r round(mean(pull(filter(posture,sex=="female"),fhp_size_mm), na.rm=TRUE), digits=2)`mm with a standard deviation of `r round(sd(pull(filter(posture,sex=="female"),fhp_size_mm), na.rm=TRUE),digits=2)`mm. While the mean male FHP size was `r round(mean(pull(filter(posture,sex=="male"),fhp_size_mm), na.rm=TRUE), digits=2)`mm, with a standard deviation of `r round(sd(pull(filter(posture,sex=="male"),fhp_size_mm),na.rm=TRUE), digits=2)`mm.


```{r, echo=FALSE, warning=FALSE}
fig_3=
posture %>% 
  ggplot(aes(x = age_group, y = fhp_size_mm, na.rm=TRUE)) + geom_boxplot(aes(fill=sex)) +   
  labs(title = "Distribution of FHP size by sex, age group", 
    x = "Age Group", 
    y = "FHP size in mm")

fig_3
```

People 18-30 appear to have higher rates of EEOP and men showed higher rates of EEOP than women across age groups. 


```{r, echo = FALSE, warning=FALSE}
fig_4=
posture %>% 
  mutate(eeop = as.numeric(ifelse(eop_size_mm > 10, "1", "0"))) %>% 
  group_by(age_group, sex) %>% 
  summarize(rate = mean(eeop)) %>% 
ggplot(aes(x=age_group, y=rate, group = sex, color = sex)) + geom_point(size=2) + geom_line(size=1) +
  labs(title = "Rate of enlarged EOP by sex, age group", 
    x = "Age Group", 
    y = "Rate of EEOP")

fig_4
```

```{r, include=FALSE, warning=FALSE}
grid.arrange(fig_3, fig_4, nrow = 1)
```

Our analysis examining at the relationship between EOP and FHP also confirms that EOP size is higher among those 18-30, and that FHP is highest for those over 60.

```{r, echo=FALSE, warning=FALSE}

posture %>% 
ggplot(aes(x = eop_size_mm, y = fhp_size_mm, na.rm=TRUE)) + 
  geom_point() + 
  labs(title = "EOP size and FHP size interaction", 
    x = "EOP size mm", 
    y = "FHP size mm")

posture %>% 
ggplot(aes(x = eop_size_mm, y = fhp_size_mm, color = age_group, na.rm=TRUE)) + 
  geom_point() + facet_grid(sex ~ age_group) +
  labs(title = "EOP size and FHP size interaction, by age, sex", 
    x = "EOP size mm", 
    y = "FHP size mm")
```

Based on the frequency, FHP greater than 40mm appears to be more prevalent in those over 60.

```{r, echo=FALSE}
posture %>% 
  filter(fhp_size_mm > 40) %>% 
  group_by(age_group) %>% 
  summarize(n_obs=n()) %>% 
  ggplot(aes(x = age_group, y = n_obs)) + geom_bar(stat = "identity", color="black", fill="salmon", position=position_dodge()) + 
  geom_text(aes(label = n_obs), vjust=1.6) +
    labs(title = "Count of FHP >40 by age group", 
    x = "Age Group", 
    y = "Count FHP >40")
```

In summary our analysis finds that there is a higher rate of EEOP in the 18-30 group and a higher rate of FHP in those over 60.


